---
title: "Chapter 1: Foundations"
author: "Shravan Vasishth (vasishth.github.io)"
date: "June 2025"
output: 
  html_document:
    theme: flatly
fontsize: 12pt
geometry: "a4paper,right=8.5cm,left=1cm,bottom=2cm,top=2cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, results='asis'}
cat('
<style>
body {
  margin-right: 120px;
}
</style>
')
```

\Large

\newpage

# Textbook

Introduction to Bayesian Data Analysis for Cognitive Science

Nicenboim, Schad, Vasishth

- Online version: https://bruno.nicenboim.me/bayescogsci/
- Source code: https://github.com/bnicenboim/bayescogsci
- Physical book: [here](https://www.routledge.com/Introduction-to-Bayesian-Data-Analysis-for-Cognitive-Science/Nicenboim-Schad-Vasishth/p/book/9780367359331)

**Be sure to read the textbook's chapter 1 in addition to watching this lecture**.

# Introduction: Motivation for this lecture

- Whenever we collect data, an implicit assumption is that the data are being generated from a **random variable**.
- Understanding the basic properties of random variables is of key importance when learning statistical modeling.
- The ideas and concepts in this lecture are often not taught in statistics courses in linguistics and psychology.
- The commonly used cookbook approach to teaching statistics leads to all kinds of misunderstandings that
  have a snowball effect and are a big part of the cause for the replication crisis and other problems in inference 
  that we see so often in empirical work in linguistics and psychology.  

It only takes about a day to understand these materials, but the content here will positively 
impact your ability to carry out statistical modeling and data analysis. 

# Discrete random variables

A random variable $X$ is a function $X : \Omega \rightarrow \mathbb{R}$ that associates to each **outcome** $\omega \in \Omega$ exactly one number $X(\omega) = x$.

$S_X$ is all the $x$'s (all the possible values of X, the **support of X**). I.e., $x \in S_X$.

## An example of a discrete RV

An example of a discrete random variable: keep tossing a coin again and again until you get a Heads.

- $X: \omega \rightarrow x$
- $\omega$: H, TH, TTH,\dots (infinite)
- $X(H)=1, X(TH)=2, X(TTH)=3$, \dots  
- $x=1,2,\dots; x \in S_X$

A second example of a discrete random variable: tossing a coin once.

- $X: \omega \rightarrow x$
- $\omega$: H, T
- $X(T)=0, X(H)=1$
- $x=0,1; x \in S_X$

### The probability mass function (PMF)

Every discrete (continuous) random variable X has associated with it a 
**probability mass (density)  function (PMF, PDF)**. 

- PMF is used for discrete distributions and PDF for continuous. 
- (Some books use PDF for both discrete and continuous distributions.)

Thinking just about discrete random variables for now:

\begin{equation}
p_X : S_X \rightarrow [0, 1] 
\end{equation}

defined by

\begin{equation}
p_X(x) = \hbox{Prob}(X(\omega) = x), x \in S_X
 \end{equation}

Example of a PMF: a random variable $X$ representing tossing a coin once.

- In the case of a fair coin, $x$ can be $0$ or $1$, and the probability of each possible event (each event is a subset of the set of possible outcomes) is $0.5$.
- Formally: $p_X(x) = \hbox{Prob}(X(\omega) = x), x \in S_X$
- The probability mass function defines the probability of each event:
$p_X(0) = p_X(1) =0.5$.

### The cumulative distribution function (CDF)

The \textbf{cumulative distribution function} (CDF) $F(X\leq x)$ gives the cumulative probability of observing all the events $X\leq x$. 

\begin{equation}
\begin{split}
F(x=1) & = \hbox{Prob}(X \leq 1)\\
       & = \sum_{x=0}^{1} p_X(x) \\
       & = p_X(x=0) + p_X(x=1) \\
       & =  1
\end{split}
\end{equation}

\begin{equation}
\begin{split}
F(x=0) & = \hbox{Prob}(X \leq 0)\\
& = \sum_{x=0}^{0} p_X(x)\\
& = p_X(x=0) \\
& = 0.5
\end{split}
\end{equation}

Do 10 coin-tossing experiments, each with one trial. The probability (which I call $\theta$ below) of heads 0.5:

```{r}
extraDistr::rbern(n = 10, prob = 0.5)
```

The probability mass function: Bernoulli 

$$p_X(x)= \theta^x (1-\theta)^{(1-x)}$$

where x can have values 0, 1.

What's the probability of a tails/heads? The d-family of functions:

```{r}
extraDistr::dbern(0, prob = 0.5)
extraDistr::dbern(1, prob = 0.5)
```

Notice that these probabilities sum to 1.


The cumulative probability distribution function: the p-family of functions: 

$$F(x=1) = Prob(X \leq 1) = \sum_{x=0}^{1} p_X(x) = 1$$

```{r}
extraDistr::pbern(1, prob = 0.5)
```

$$F(x=0) = Prob(X \leq 0) = \sum_{x=0}^{0} p_X(x) = 0.5$$

```{r}
extraDistr::pbern(0, prob = 0.5)
```

## Another example of a discrete random variable: The binomial

- Consider carrying out a single experiment where you toss a coin 10 times (the number of trials, `size` in R). 
- When the number of trials (size) is 1, we have a Bernoulli; when we have size greater than 1, we have a Binomial.


$$\theta^{x}(1-\theta)^{1-x}$$ where $$S_X=\{0,1\}$$

**Binomial PMF**

$${n \choose x} \theta^{x}(1-\theta)^{n-x}$$ 

where $$S_X=\{0,1,\dots,n\}$$

- $n$ is the number of times the coin was tossed (the number of trials; size in R).
- ${n \choose x}$ is the number of ways that you can get $x$ successes in $n$ trials.

```{r}
choose(10, 2)
```
- $\theta$ is the probability of success in $n$ trials.


```{r echo=FALSE,message=FALSE}
library(ggplot2)
library(tidyverse)
ggplot(tibble(x = c(0, 10)), aes(x)) +
  stat_function(
    fun = dbinom, geom = "col",
    args = list(size = 10, prob = 0.5), n = 11
  ) +
  xlab("Possible events") +
  ylab("Probability") +
  ggtitle(expression(paste(theta, " = 0.5", sep = ""))) +
  scale_x_continuous(breaks = 0:10) +
  theme_bw()
```

```{r echo=FALSE,message=FALSE}
ggplot(tibble(x = c(0, 10)), aes(x)) +
  stat_function(
    fun = dbinom, geom = "col",
    args = list(size = 10, prob = 0.1), n = 11
  ) +
  xlab("Possible events") +
  ylab("Probability") +
  ggtitle(expression(paste(theta, " = 0.1", sep = ""))) +
  scale_x_continuous(breaks = 0:10) +
  theme_bw()
```

```{r echo=FALSE,message=FALSE}
ggplot(tibble(x = c(0, 10)), aes(x)) +
  stat_function(
    fun = dbinom, geom = "col",
    args = list(size = 10, prob = 0.9), n = 11
  ) +
  xlab("Possible events") +
  ylab("Probability") +
  ggtitle(expression(paste(theta, " = 0.9", sep = ""))) +
  scale_x_continuous(breaks = 0:10) +
  theme_bw()
```

## Four critical R functions for the binomial RV

**1. Generate random data: rbinom**

- n: number of experiments done (**Note**: in the binomial pdf, n stands for the number of trials). In R, n is 
called the number of observations.
- size: the number of times the coin was tossed in each experiment (the number of trials)

Example: 10 separate experiments, each with 1 trial:

```{r}
rbinom(n = 10, size = 1, prob = 0.5)
## equivalent to: rbern(10,0.5)
```

Example: 10 separate experiments, each with 10 trials: 

```{r}
rbinom(n = 10, size = 10, prob = 0.5)
```

**2. Compute probabilities of particular events (0,1,...,10 successes when n=10): dbinom**

```{r}
probs <- round(dbinom(0:10, size = 10, 
                      prob = 0.5), 3)
x <- 0:10
```

```{r echo=FALSE}
(probs_df<-data.frame(x,probs))
```

**3. Compute cumulative probabilities: pbinom**

```{r echo=FALSE}
#round(pbinom(0:10, size = 10, prob = 0.5),3)
```


```{r echo=FALSE,message=FALSE}
ggplot(tibble(q = c(0, 10)), aes(q)) +
  stat_function(
    fun = pbinom, geom = "col",
    args = list(size = 10, prob = 0.5), n = 11
  ) +
  xlab("Possible events k") +
  ylab("Prob(X<=k)") +
  ggtitle("Cumulative distribution function") +
  scale_x_continuous(breaks = 0:10) +
  theme_bw()
```

**4. Compute quantiles using the inverse of the CDF: qbinom**

```{r}
probs <- pbinom(0:10, size = 10, prob = 0.5)
qbinom(probs, size = 10, prob = 0.5)
```


```{r echo=FALSE,message=FALSE}
probs <- seq(0, 1, by = 0.01)
invcdf <- data.frame(
  probs = probs,
  q = qbinom(probs,
    size = 10,
    prob = 0.5
  )
)

ggplot(invcdf, aes(
  x = probs,
  y = q
)) +
  geom_line() +
  xlim(0, 1) +
  xlab("probability") +
  ylab("quantile") +
  ggtitle("Inverse CDF,  binomial(size=10,prob=0.5)") +
  scale_y_continuous(breaks = seq(0, 10, by = 1)) +
  theme_bw()
```

These four functions are the d-p-q-r family of functions, and are available for all
the distributions available in R (e.g., Poisson, geometric, normal, beta, uniform, gamma, exponential, Cauchy,  etc.).

# Continuous random variables

In coin tosses, H and T are discrete possible outcomes.

- By contrast, variables like reading times range from 0 milliseconds up--these are **continuous variables**.
- Continuous random variables have a probability **density** function (PDF) $f(\cdot)$ associated with them. (cf. PMF in discrete RVs)
- The expression 

\begin{equation}
 X \sim f(\cdot)
\end{equation}

means that the random variable $X$ is assumed to have PDF $f(\cdot)$. 

For example, if we say that $X\sim Normal(\mu,\sigma)$, we are assuming that the PDF is

\begin{equation}
f(x\mid \mu, \sigma)= \frac{1}{\sqrt{2\pi \sigma^2}} \exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right]
\end{equation}

where $-\infty<x<+\infty$

We can **truncate** the normal distribution such that $S_X$ is bounded between some lower bound and/or upper bound--this comes later.

## The normal random variable

The PDF below is associated with the normal distribution that you are probably familiar with:

\begin{equation}
f(x\mid \mu, \sigma)= \frac{1}{\sqrt{2\pi \sigma^2}} \exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right] 
\end{equation}

where  $-\infty<x<+\infty$.

- The support of $X$, i.e., the elements of $S_X$, has values ranging from $-\infty$ to $+\infty$
- $\mu$ is the location parameter (here, mean)
- $\sigma$ is the scale parameter (here, standard deviation)


In the discrete RV case, we could compute the probability of a \textbf{particular} event occurring:

```{r}
extraDistr::dbern(x = 1, prob = 0.5)
dbinom(x = 2, size = 10, prob = 0.5)
```

- In a continuous distribution, probability is defined as the **area under the curve**.
- As a consequence, for any particular **point** value $x$, where $X\sim Normal(\mu,\sigma)$, it is always the case that $\hbox{Prob}(X=x) = 0$.
- In any continuous distribution, we can compute probabilities like $\hbox{Prob}(x_1<X<x_2) = ?$, where $x_1<x_2$, by summing up the **area under the curve**.
- To compute probabilities like $\hbox{Prob}(x_1<X<x_2) = ?$, we need the cumulative distribution function.

The cumulative distribution function (CDF) is

\begin{equation}
P(X<u) = F(X<u) =\int_{-\infty}^{u} f(x)\, dx
\end{equation}

- The integral sign $\int$ is just the summation symbol in continuous space.
- Recall the summation in the CDF of the Bernoulli!

## The standard normal distribution

In the $Normal(\mu=0,\sigma=1)$,

- Prob$(-1<X<+1) = 0.68$
- Prob$(-2<X<+2) = 0.95$
- Prob$(-3<X<+3) = 0.997$

```{r echo=FALSE}
# draw the normal curve
curve(dnorm(x, 0, 1), xlim = c(-3, 3), main = "Normal density")

# define shaded region
from.z <- -1
to.z <- 1

S.x <- c(from.z, seq(from.z, to.z, 0.01), to.z)
S.y <- c(0, dnorm(seq(from.z, to.z, 0.01)), 0)
polygon(S.x, S.y, col = rgb(1, 0, 0, 0.3))
text(-0.7, 0.15, pos = 4, cex = 3.5, expression(paste("68%")))
```

```{r, echo=FALSE}
# draw the normal curve
# draw the normal curve
curve(dnorm(x, 0, 1), xlim = c(-3, 3), main = "Normal density")

# define shaded region
from.z <- -2
to.z <- 2

S.x <- c(from.z, seq(from.z, to.z, 0.01), to.z)
S.y <- c(0, dnorm(seq(from.z, to.z, 0.01)), 0)
polygon(S.x, S.y, col = rgb(0, 0, 1, 0.3))
text(-0.7, 0.15, pos = 4, cex = 3.5, expression(paste("95%")))
```

```{r echo=FALSE}
# draw the normal curve
# draw the normal curve
curve(dnorm(x, 0, 1), xlim = c(-4, 4), main = "Normal density")

# define shaded region
from.z <- -3
to.z <- 3

S.x <- c(from.z, seq(from.z, to.z, 0.01), to.z)
S.y <- c(0, dnorm(seq(from.z, to.z, 0.01)), 0)
polygon(S.x, S.y, col = rgb(0, 0, 0, 0.3))
text(-1, 0.15, pos = 4, cex = 3.5, expression(paste("99.7%")))
```

More generally, for any $Normal(\mu,\sigma)$,

- Prob$(-1\times \sigma <X<+1\times \sigma) = 0.68$
- Prob$(-2\times \sigma <X<+2\times \sigma) = 0.95$
- Prob$(-3\times \sigma<X<+3\times \sigma) = 0.997$

## The normalizing constant and the kernel

The PDF of the normal again:

\begin{equation}
f(x\mid \mu, \sigma)= \frac{1}{\sqrt{2\pi \sigma^2}} \exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right] 
\end{equation}


This part of $f(x\mid \mu,\sigma)$ (call it $g(x)$) is  the ``kernel'' of the normal PDF:

\begin{equation}
g(x\mid \mu, \sigma) = \exp\left[-\frac{(x-\mu)^2}{2 \sigma^2}\right]
\end{equation}

For the above function, the area under the curve doesn't sum to 1:

Sum up the area under the curve $\int g(x)\, dx$:

```{r}
g <- function(x, mu = 0, sigma = 1) {
  exp((-(x - mu)^2 / (2 * (sigma^2))))
}

integrate(g, lower = -Inf, upper = +Inf)$value
```

The shape doesn't change of course:

```{r echo=FALSE}
x <- seq(-10, 10, by = 0.01)

plot(function(x) g(x), -3, 3,
  main = "Normal density kernel", ylim = c(0, 1),
  ylab = "density", xlab = "X"
)
```

```{r echo=FALSE}
x <- seq(-10, 10, by = 0.01)

plot(function(x) dnorm(x), -3, 3,
  main = "Normal density", ylim = c(0, 0.5),
  ylab = "density", xlab = "X"
)
```

In simple examples like the one shown here, given the kernel of some PDF like $g(x)$, we can figure out the normalizing constant by solving for $k$ in:

\begin{equation}
k \int g(x)\, dx = 1
\end{equation}

Solving for $k$ just amounts to computing:

\begin{equation}
k  = \frac{1}{\int g(x)\, dx}
\end{equation}

So, in our example above, 

```{r}
(k<-1/integrate(g, lower = -Inf, upper = +Inf)$value)
```

The above number is just $\frac{1}{\sqrt{2 \pi \sigma^2}}$, where $\sigma=1$:

```{r}
1/(sqrt(2*pi*1))
```

Once we include the normalizing constant, the area under the curve in $g(x)$ sums to 1:

```{r}
k * integrate(g, lower = -Inf, upper = +Inf)$value
```


We will see the practical implication of this when we move on to chapter 2 of the textbook.

## The d-p-q-r functions for the normal distribution

In the continuous case, we also have this family of d-p-q-r functions. In the normal distribution:

**1. Generate random data using rnorm**

```{r}
round(rnorm(5, mean = 0, sd = 1),3)
```

For the standard normal, mean=0, and sd=1 can be omitted (these are the default values in R).

```{r}
round(rnorm(5),3)
```


**2. Compute probabilities using CDF: pnorm**

Some examples of usage:

- $\hbox{Prob}(X<2)$ (e.g., in $X\sim Normal(0,1)$)

```{r}
pnorm(2)
```

- $\hbox{Prob}(X>2)$ (e.g., in $X\sim Normal(0,1)$)

```{r}
pnorm(2, lower.tail = FALSE)
```



**3. Compute quantiles: qnorm**

```{r}
qnorm(0.9772499)
```


**4. Compute the probability density: dnorm**

```{r}
dnorm(2)
```

**Note**: In the continuous case, this is a **density**, the value $f(x)$, not a probability. Cf. the discrete examples dbern and dbinom, which give probabilities of a point value  $x$.

```{r echo=FALSE}
curve(dnorm(x, 0, 1),
  xlim = c(-3, 3), main = "Normal(0,1)",
  ylab = "density"
)
from.z <- -1
to.z <- 1

S.x <- c(from.z, seq(from.z, to.z, 0.01), to.z)
S.y <- c(0, dnorm(seq(from.z, to.z, 0.01)), 0)
polygon(S.x, S.y, col = rgb(1, 0, 0, 0.3))
text(-2, 0.15, pos = 4, cex = 1.5, paste("pnorm(1)-pnorm(-1)"))
arrows(x1 = 2, y1 = 0.3, x0 = 1, y0 = dnorm(1), code = 1)
text(1.7, 0.32, pos = 4, cex = 1.5, paste("dnorm(1)"))
points(1, dnorm(1))
# points(1,0)
arrows(x1 = 2, y1 = 0.1, x0 = 1, y0 = 0, code = 1)
text(1, 0.12, pos = 4, cex = 1.5, paste("qnorm(0.841)"))
x <- rnorm(10)
points(x = x, y = rep(0, 10), pch = 17)
text(-3, 0.05, pos = 4, cex = 1.5, paste("rnorm(10)"))
arrows(x1 = -2.5, y1 = 0.03, x0 = min(x), y0 = 0, code = 1)
```

```{r echo=FALSE}
ggplot(tibble(x = c(200, 900)), aes(x)) +
  stat_function(
    fun = dnorm,
    args = list(mean = 500, sd = 100)
  ) +
  xlab("x") +
  ylab("density") +
  ggtitle("PDF of Normal(mean=500,sd=100)") +
  theme_bw()
```

```{r echo=FALSE}
ggplot(tibble(x = c(200, 900)), aes(x)) +
  stat_function(
    fun = pnorm,
    args = list(mean = 500, sd = 100)
  ) +
  xlab("x") +
  ylab("Probability") +
  ggtitle("CDF of Normal(mean=500,sd=100)") +
  theme_bw()
```


```{r echo=FALSE}
ggplot(tibble(y = c(0, 1)), aes(y)) +
  stat_function(
    fun = qnorm,
    args = list(mean = 500, sd = 100)
  ) +
  xlab("Probability") +
  ylab("x") +
  ggtitle("Inverse CDF of Normal(mean=500,sd=100)") +
  theme_bw()
```


# The likelihood function (Binomial)

The **likelihood function** refers to the PMF  $p(k|n,\theta)$, treated as a function of $\theta$. 

For example, suppose that we record $n=10$ trials, and observe $k=7$ successes. The likelihood function is:

\begin{equation}
\mathcal{L}(\theta|k=7,n=10)=
\binom{10}{7} \theta^{7} (1-\theta)^{10-7}
\end{equation}

If we now plot the likelihood function for all possible values of $\theta$ ranging from $0$ to $1$, we get the plot shown below.

```{r echo=FALSE}
ggplot(tibble(prob = c(0, 1)), aes(prob)) +
  stat_function(
    fun = dbinom,
    args = list(x = 7, size = 10)
  ) +
  xlab("theta") +
  ylab("Likelihood") +
  geom_vline(xintercept = .7, linetype = "dashed") +
  annotate("text", x = .55, y = .05, label = "Max. value at: \n 0.7") +
  ggtitle("Likelihood function") +
  theme_bw()
```

The MLE (**from a particular sample** of data need not invariably give us an accurate estimate of $\theta$. 

```{r echo=FALSE}
n <- c(seq(10, 100, by = 1), seq(100, 1000, by = 10), seq(1000, 10000, by = 50), seq(10000, 100000, by = 100))
mean_store <- rep(NA, length(n))
for (i in 1:length(n)) {
  y <- rbinom(n[i], size = 1, prob = 0.7)
  mean_store[i] <- mean(y)
}

mle <- data.frame(n = n, mean_store = mean_store)

ggplot(mle, aes(x = n, y = mean_store)) +
  # geom_line() +
  geom_point() +
  scale_x_log10() +
  ylim(0, 1) +
  xlab("sample size n (log scale)") +
  ylab("estimate") +
  geom_hline(yintercept = 0.7) +
  ggtitle("The MLE  as a function of sample size (funnel plot)") +
  theme_bw()
```

Sample size is key here: as $n\rightarrow \infty$, we approach the true value of the parameter (here, $\theta$).

# The likelihood function (Normal)


\begin{equation}
\mathcal{L}(\mu,\sigma|x)=Normal(x,\mu,\sigma)
\end{equation}

Below, assume that $\sigma=1$.

```{r}
## the data:
x<-0
## the likelihood under different values 
## of mu:
dnorm(x,mean=0,sd=1)
dnorm(x,mean=10,sd=1)
```

Assuming that $\sigma=1$, the likelihood function of $\mu$:

```{r echo=FALSE}
x<-0
muvals<-seq(-4,4,by=0.01)
plot(muvals,dnorm(0,mean=muvals,sd=1),
     type="l",ylab="Density")
abline(v=0)
```

If we have two **independent** data points, the joint likelihood given the data of $\mu$, assuming $\sigma=1$:

```{r}
x1<-0
x2<-1.5
dnorm(x1,mean=0,sd=1) * 
  dnorm(x2,mean=0,sd=1) 
## log likelihood:
dnorm(x1,mean=0,sd=1,log=TRUE) + 
  dnorm(x2,mean=0,sd=1,log=TRUE) 
## more compactly:
x<-c(x1,x2)
sum(dnorm(x,mean=0,sd=1,log=TRUE))
```

One practical implication: one can use the log likelihood to 
compare competing models' fit:

```{r}
## Model 1:
sum(dnorm(x,mean=0,sd=1,log=TRUE))
## Model 2:
sum(dnorm(x,mean=10,sd=1,log=TRUE))
```

Model 1 has higher likelihood than Model 2, so we'd prefer to assume that the data are better characterized by Model 1 than 2 (neither may be the true model!).

More generally, for independent and identically distributed data $x=x_1,\dots,x_n$:

\begin{equation}
\mathcal{L}(\mu,\sigma|x) = \prod_{i=1}^n Normal(x_i,\mu,\sigma)
\end{equation}

or

\begin{equation}
\ell (\mu,\sigma|x) = \sum_{i=1}^n log(Normal(x_i,\mu,\sigma))
\end{equation}

# The expectation and variance of an RV

Read section 1.4.1 of chapter 1 of the textbook, and (optionally) chapter 2 of the linear modeling lecture notes here: 

https://github.com/vasishth/LM

# Bivariate/multivariate distributions

## Discrete bivariate distributions

```{r echo=FALSE}
library(bcogsci)
```

Data from: Laurinavichyute, A. (2020). Similarity-based interference and faulty encoding accounts of sentence processing. dissertation, University of Potsdam.

X: Likert ratings 1-7.

Y: 0, 1 accuracy responses.

```{r echo=FALSE}
data("df_discreteagrmt")
rating0 <- df_discreteagrmt %>%
  filter(accuracy == 0) %>%
  count(rating) %>%
  pull(n)
rating1 <- df_discreteagrmt %>%
  filter(accuracy == 1) %>%
  count(rating) %>%
  pull(n)
ratingsbivar <- tibble(`0` = rating0, `1` = rating1)
## function from the bivariate package:
bivariate::gbvpmf(as.matrix(ratingsbivar)) %>%
  plot(arrows = FALSE)

rating0 <- df_discreteagrmt %>%
  filter(accuracy == 0) %>%
  count(rating) %>%
  pull(n)
rating1 <- df_discreteagrmt %>%
  filter(accuracy == 1) %>%
  count(rating) %>%
  pull(n)
ratingsbivar <- tibble(`0` = rating0, `1` = rating1)
## function from the bivariate package:
bivariate::gbvpmf(as.matrix(ratingsbivar)) %>%
  plot(arrows = FALSE)

probs <- t(attr(bivariate::gbvpmf(as.matrix(ratingsbivar)), "p"))
colnames(probs) <- paste("x=", 1:7, sep = "")
rownames(probs) <- paste("y=", 0:1, sep = "")
```

The joint PMF: $p_{X,Y}(x,y)$

\small

```{r echo=FALSE}
probs_f <- probs %>%
  apply(2, function(.x) paste0("$", round(.x, 3), "$"))
colnames(probs_f) <- paste0("$", colnames(probs), "$")
rownames(probs_f) <- paste0("$", rownames(probs), "$")
kableExtra::kable(probs_f,
  digits = 3,
  escape = FALSE,
  format = "latex",
  caption = "The joint PMF for two random variables X and Y."
)
```


\Large
For each possible pair of values of X and Y, we have a **joint probability mass function** $p_{X,Y}(x,y)$.

Two useful quantities that we can compute:

### The marginal distributions ($p_{X}$ and $p_Y$)

\begin{equation}
p_{X}(x)=\sum_{y\in S_{Y}}p_{X,Y}(x,y).\label{eq-marginal-pmf2}
\end{equation}	

\begin{equation}
p_{Y}(y)=\sum_{x\in S_{X}}p_{X,Y}(x,y).\label{eq-marginal-pmf}
\end{equation}	


\small

```{r echo=FALSE}
# P(Y)
PY <- rowSums(probs)
# P(X)
PX <- colSums(probs)
probs_f <- probs %>%
  apply(2, function(.x) paste0("$", round(.x, 3), "$"))

## probs <- round(probs,3)
## rownames(probs)[3]<-"P(X)"
## colnames(probs)[8]<-"P(Y)"

probs_f <- rbind(probs_f, paste0("$", round(PX, 3), "$"))
probs_f <- cbind(probs_f, c(paste0("$", round(PY, 3), "$"), ""))
colnames(probs_f) <- paste0("$", c(colnames(probs), "p(Y)"), "$")
rownames(probs_f) <- paste0("$", c(rownames(probs), "p(X)"), "$")

kableExtra::kable(probs_f,
  escape = FALSE,
  digits = 3, booktabs = TRUE,
  vline = "", # format="latex",
  caption = "The joint PMF for two random variables X and Y, along with the marginal distributions of X and Y."
)
```  

\Large


```{r echo=FALSE}
op <- par(mfrow = c(1, 2), pty = "s")
barplot(PX, main = "p(X)")
barplot(PY, main = "p(Y)")
```

### The conditional distributions ($p_{X|Y}$ and $p_{Y|X}$)

\begin{equation}
p_{X\mid Y}(x\mid y) = \frac{p_{X,Y}(x,y)}{p_Y(y)}	
\end{equation}

and 

\begin{equation}
p_{Y\mid X}(y\mid x) = \frac{p_{X,Y}(x,y)}{p_X(x)}	
\end{equation}


Let's do the calculation for $p_{X\mid Y}(x\mid y=0)$.

\small

```{r echo=FALSE}
kableExtra::kable(probs_f,
  escape = FALSE,
  digits = 3, booktabs = TRUE,
  vline = "", # format="latex",
  caption = "The joint PMF for two random variables X and Y, along with the marginal distributions of X and Y."
)
```

\Large

\begin{equation}
\begin{split}
p_{X\mid Y} (1\mid 0) =& \frac{p_{X,Y}(1,0)}{p_Y(0)}\\
	=&  \frac{0.018}{0.291}\\
	=& `r round(0.018/0.291,3)`
\end{split}	
\end{equation}

As an exercise, figure out the conditional distribution of X given Y, and the conditional distribution of Y given X. 


## Continuous bivariate distributions

Next, we turn to continuous bivariate/multivariate distributions.

The variance-covariance matrix:

\begin{equation}\label{eq:covmatfoundations}
\Sigma
=
\begin{pmatrix}
\sigma _{X}^2  & \rho_{XY}\sigma _{X}\sigma _{Y}\\
\rho_{XY}\sigma _{X}\sigma _{Y}    & \sigma _{Y}^2\\
\end{pmatrix}
\end{equation}

The off-diagonals of this matrix contain the covariance between  $X$ and $Y$:

$$Cov(X,Y) = \rho_{XY}\sigma _{X}\sigma _{Y}$$ 

The joint distribution of $X$ and $Y$ is defined as follows:

\begin{equation}\label{eq:jointpriordistfoundations}
\begin{pmatrix}
  X \\ 
  Y \\
\end{pmatrix}
\sim 
\mathcal{N}_2 \left(
\begin{pmatrix}
  0 \\
  0 \\
\end{pmatrix},
\Sigma
\right)
\end{equation}

The joint PDF has the property that the volume under the surface sums to 1.

Formally, we would write the volume under the surface as a double integral: we are summing up the volume under the surface for both $X$ and $Y$ (hence the two integrals).

\begin{equation}
\iint_{S_{X,Y}} f_{X,Y}(x,y)\, dx dy = 1
\end{equation}


Here, the terms $dx$ and $dy$ express the fact that we are computing the volume under the surface along the $X$ axis and the $Y$ axis.

The joint CDF would be written as follows. The equation below gives us the probability of observing a value like $(u,v)$ or some value smaller than that (i.e., some $(u',v')$, such that $u'<u$ and $v'<v$).

\begin{equation}
\begin{split}
F_{X,Y}(u,v) =& \hbox{Prob}(X<u,Y<v)\\
             =& \int_{-\infty}^u \int_{-\infty}^v f_{X,Y}(x,y)\, dy dx \\
             ~& \hbox{ for } (x,y) \in \mathbb{R}^2\\
\end{split}
\end{equation}


Just as in the discrete case, the marginal distributions can be derived by marginalizing out the other random variable:

\begin{equation}
f_X(x) = \int_{S_Y} f_{X,Y}(x,y)\, dy \quad f_Y(y) = \int_{S_X} f_{X,Y}(x,y)\, dx
\end{equation}

Here, $S_X$ and $S_Y$ are the respective supports.

```{r echo=FALSE, fig.caption="No correlation"}
f1 <- bivariate::nbvpdf(0, 0, 1, 1, 0)
# f1 %$% matrix.variances
plot(f1,
  all = TRUE, n = 20,
  main = "No correlation"
)
```

```{r echo=FALSE}
f3 <- bivariate::nbvpdf(0, 0, 1, 1, -0.6)
# f1 %$% matrix.variances
plot(f3,
  all = TRUE, n = 20,
  main = "Correlation -0.6"
)
```

```{r echo=FALSE}
f2 <- bivariate::nbvpdf(0, 0, 1, 1, 0.6)
# f1 %$% matrix.variances
plot(f2,
  all = TRUE, n = 20,
  main = "Correlation 0.6"
)
```


## Generate simulated bivariate (multivariate) data 

```{r}
## define a variance-covariance matrix:
Sigma <- matrix(c(5^2, 5 * 10 * 0.6, 
                  5 * 10 * 0.6, 10^2),
  byrow = FALSE, ncol = 2
)
## generate data:
u <- MASS::mvrnorm(n = 100, mu = c(0, 0), 
                   Sigma = Sigma)
head(u, n = 3)
```


```{r echo=FALSE}
ggplot(tibble(u_1 = u[, 1], u_2 = u[, 2]), 
aes(u_1, u_2)) +
  geom_point() +
  theme_bw()
```

One practical implication: Such bi/multivariate distributions
become critically important to understand when we turn to  hierarchical (linear mixed) models.